---
title: "PGR Training: Bayesian modelling in R and Stan<br> Worksheet"
author: "Matteo Lisi"
format:
  html:
    toc: true
    code-tools: true     # run/copy buttons
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(rstan)
library(lme4)
```



## Activity 1 — MCMC "by hand" with Metropolis algorithm (simple linear regression)

> Goal: Implement a bare-bones Metropolis sampler for a straight-line model on one participant from `sleepstudy`. Inspect priors, define the log-likelihood and (unnormalized) log-posterior, write a simple random-walk proposal, run the chain, and compare to least-squares.

#### Setup

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 4
#| fig-height: 4.5

# Packages used here
library(lme4) # for the sleepstudy data

# alternatively, if your don't have lme4 installed
sleepstudy <- read.csv("data/sleepstudy.csv")

# Pick example subject and plot their data
par(mfrow = c(1,1))
with(sleepstudy[sleepstudy$Subject == "308",],
plot(Days, Reaction, pch = 19, main = "Subject 308",
xlab = "Days of sleep deprivation", ylab = "Reaction time (ms)",
xlim = c(-1, 11), ylim = c(200, 500)))


# Save vectors for the model code below
x <- sleepstudy$Days[ sleepstudy$Subject == "308" ]
y <- sleepstudy$Reaction[ sleepstudy$Subject == "308" ]
```

##### 1) Log-likelihood

The model is:$$y_i \sim \mathcal{N}(\alpha + \beta x_i, \sigma_{\epsilon}^2)$$

Where $\alpha$ is the intercept, and $\beta$ the slope. We will work with $\log \sigma_{\epsilon}$ as the third parameter for stability (and so that we don't need to worry about the lower boundary at zero).

```{r}
# par = c(intercept, slope, log_sigma)
loglik <- function(par){
  pred <- par[1] + par[2] * x
  sum(dnorm(y, mean = pred, sd = exp(par[3]), log = TRUE))
}
```

##### 2) Priors (and a quick visual check)

We use an informative prior, as there is relevant prior knowledge that we can rely on (e.g. that average simple reaction times in humans tend to be around 250 ms):
- $\alpha \sim \mathcal{N} (250, 180^2)$
- $\beta \sim \mathcal{N} (20, 20^2)$
- $\log \sigma_{\epsilon} \sim \mathcal{N} (4, 1^2)$


```{r}
logprior <- function(par){
  dnorm(par[1], mean = 250, sd = 180, log = TRUE) +
  dnorm(par[2], mean = 20, sd = 20, log = TRUE) +
  dnorm(par[3], mean = 4, sd = 1, log = TRUE)
}
```

Is it useful to visualise it: these are our _a prior_ expectations about plausible parameter values.

```{r}
#| fig-align: 'center'
#| fig-width: 9
#| fig-height: 3


par(op <- par(no.readonly = TRUE)); on.exit(par(op), add = TRUE)
par(mfrow = c(1,3), mar = c(4,4,2,1))
curve(dnorm(x, mean = 250, sd = 180), from = 0, to = 1000,
xlab = "Intercept", ylab = "prior density", col = "blue")
curve(dnorm(x, mean = 20, sd = 20), from = -50, to = 50,
xlab = "Slope", ylab = "prior density", col = "blue")
xx <- seq(-1, 6, length.out = 500)
yy <- dnorm(xx, mean = 4, sd = 1)
plot(exp(xx), yy, type = "l", xlab = expression(sigma[epsilon]),
ylab = "prior density", col = "blue")

```

##### 3) Unnormalized log-posterior

```{r}
logposterior <- function(par){
loglik(par) + logprior(par)
}
```


##### 4) Random-walk proposal

We use a diagonal Gaussian proposal with tunable step sizes.

```{r}
proposal_sd <- c(15, 5, 0.2) # try tweaking these later
  proposalfunction <- function(par){
  rnorm(3, mean = par, sd = proposal_sd)
}
```


##### 5) Metropolis sampler

```{r}
run_metropolis_MCMC <- function(startvalue, iterations){
  
  # set up an empty array to store sampled values
  chain <- array(dim = c(iterations+1,3))
  
  # put starting values at top of arrays
  chain[1,] <- startvalue
  
  for (i in 1:iterations){
    
    # draw a random proposal
    proposal <- proposalfunction(chain[i,])
    
    # ratio of posterior density between new and old values
    a <- exp(logposterior(proposal) - logposterior(chain[i,]))
    
    # sample random number & accept/reject the parameter values
    if (runif(1) < a){
      chain[i+1,] <- proposal
    }else{
      chain[i+1,] <- chain[i,]
    }
  }
  return(chain)
}
```

##### 6) Run and visualize

```{r}
# Settings
startvalue <- c(250, 20, 5) # c(intercept, slope, log_sigma)
iterations <- 20000


# Run chain
chain <- run_metropolis_MCMC(startvalue, iterations)


# Quick diagnostics and LS reference
burnIn <- 3000
acceptance <- 1 - mean(duplicated(chain[-(1:burnIn), ]))
LSfit <- lm(y ~ x)
interceptLS <- coef(LSfit)[1]
slopeLS <- coef(LSfit)[2]
sigmaLS <- summary(LSfit)$sigma


# Plots
par(mfrow = c(2,3))
hist(chain[-(1:burnIn), 1], main = "Intercept", border = "white",
col = "dark grey", breaks = 20)
abline(v = interceptLS, col = "red", lwd = 2)


hist(chain[-(1:burnIn), 2], main = "Slope", border = "white",
col = "dark grey", breaks = 20)
abline(v = slopeLS, col = "red", lwd = 2)


hist(exp(chain[-(1:burnIn), 3]), main = expression(sigma[epsilon]),
border = "white", col = "dark grey", breaks = 20)
abline(v = sigmaLS, col = "red", lwd = 2)


plot(chain[-(1:burnIn), 1], type = "l", main = "Chain values: intercept")
abline(h = interceptLS, col = "red", lwd = 2)
plot(chain[-(1:burnIn), 2], type = "l", main = "Chain values: slope")
abline(h = slopeLS, col = "red", lwd = 2)
plot(exp(chain[-(1:burnIn), 3]), type = "l", main = "Chain values: sigma")
abline(h = sigmaLS, col = "red", lwd = 2)
```


##### 7) Summaries (credible interval for slope)

```{r}
# Remove initial burn-in
burnIn <- 5000
slope_samples <- chain[-(1:burnIn), 2]


# Posterior mean
mean(slope_samples)


# 95% credible interval (percentile method)
alpha <- 0.05
round(c(quantile(slope_samples, probs = alpha/2),
quantile(slope_samples, probs = 1 - alpha/2)), 2)
```



#### Further task

Experiment by changing the priors and the proposal distribution, then run the sampler again. Observe how these changes affect convergence, acceptance rate, and the shape and location of the posterior distributions.



## Activity 2 — Linear regression in Stan

> Goal: Fit a no-intercept Bayesian linear regression to the Hubble data in Stan, check convergence, summarize the posterior for β and σ, and then transform the posterior for β into a posterior for the age of the universe (in billions of years). Finally, visualize uncertainty both on the parameter and data scales.

#### Setup

Load data

```{r}
d <- read.csv("data/hubble.csv")
str(d)
head(d)
```


Quick plot

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 4
#| fig-height: 4.5

plot(d$distance, d$velocity,
     xlab="Distance [Mpc]", ylab="Velocity [km/s]",
     pch=19)
```

##### 1) Stan code

Saved in the repository as `stan_code/hubble_model.stan`

```{.stan}

data {
  int<lower=1> N;
  vector[N] distance;   // Mpc
  vector[N] velocity;   // km/s
}
parameters {
  real<lower=0> beta;   // (km/s)/Mpc
  real<lower=0> sigma;  // km/s
}
model {
  // weakly-informative half-normal style via <lower=0>
  beta  ~ normal(0, 200);
  sigma ~ normal(0, 2000);
  velocity ~ normal(beta * distance, sigma);
}

```



##### 2) Preparing data for modelling in Stan

We prepare data as list, and include also the number of observations `N`, which is needed in Stan to declare the size of data vectors.

```{r}
stan_data <- list(
  N = nrow(d),
  distance = d$distance,  # Mpc
  velocity = d$velocity   # km/s
)
str(stan_data)
```


##### 3) Fit the model and check convergence

```{r, echo=FALSE}
fit <- readRDS("outputs/hubble_fit.RDS")
```

Load `rstan` library and run sampling of posterior

```{r, eval=FALSE}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = max(1, parallel::detectCores() - 1))

fit <- stan(
  file = "stan_code/hubble_model.stan",
  data = stan_data,
  iter = 2000,
  chains = 4)

```


Extract the posterior samples, and store them in a database format using the `tidybayes` package


```{r}
library(tidyverse)
library(tidybayes)

posterior_samples <- fit %>%
  spread_draws(beta) 

head(posterior_samples)
```


Examine results & assess convergence

```{r}
print(fit, pars = c("beta","sigma"), probs = c(.025,.5,.975))
traceplot(fit, pars = c("beta","sigma"))
```


##### 3) Extract posterior draws and estimate the age of the universe

```{r}
mpc_km <- 3.09e19 # km per Megaparsec
sec_per_year <- 60^2 * 24 * 365

# transform in Km
hubble.const <- posterior_samples$beta/ mpc_km 

# invert to get age in seconds
age <- 1/hubble.const

# transform age in billion years
age <- (age/sec_per_year) / 10^9

# point estimate
mean(age)
```

We can use a percentile method to determine a credibility interval

```{r}
# 95% Bayesian credible interval
alpha <- 0.05
round(c(quantile(age, probs = alpha/2),
        quantile(age, probs = 1-alpha/2)),
      digits=2)
```


::: callout-tip

#### Highest posterior density interval

The percentile method works well in most cases but can be misleading for asymmetric posteriors, where for example the mode (the most probable value) may fall outside the interval.
A highest posterior density (HPD, also referred to as HDI) interval is the shortest interval containing a specified proportion of the posterior mass. _Every point inside has higher probability density than any point outside._

In our example, the posterior for β is roughly symmetric, so percentile and HPD intervals are nearly identical.

Here is how we can compute this using the handy `tidybayes` package

```{r}
fit %>%
  spread_draws(beta) %>% 
  mean_hdi(beta, .width = 0.95)
```

Or alternatively, computing the interval for more than one parameter at the same time:

```{r}
fit %>%
  gather_draws(beta, sigma) %>%  
  group_by(.variable) %>%
  mean_hdi(.value, .width = 0.95)
```


:::

##### 4) Visualisation of uncertainty

Visualise posterior over the age of the universe as simple histogram

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 6
#| fig-height: 4

hist(age, breaks =30, xlab="time (billions of years)")
```



We can make this a bit nicer e.g. by adding smooth density curve and credible interval

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 6
#| fig-height: 4

# Get HDI interval
hdi <- mean_hdi(age, .width = 0.95)

# Compute density
dens <- density(age, adjust=2)

# Plot
plot(dens, main = "", xlab = "time (billions of years)", lwd = 2)

# Fill inside HDI
polygon(c(dens$x[dens$x >= hdi$ymin & dens$x <= hdi$ymax], hdi$ymax, hdi$ymin),
        c(dens$y[dens$x >= hdi$ymin & dens$x <= hdi$ymax], 0, 0),
        col = "blue", border = NA)

# Fill outside HDI
polygon(c(dens$x[dens$x < hdi$ymin], hdi$ymin, dens$x[1]),
        c(dens$y[dens$x < hdi$ymin], 0, 0),
        col = "gray90", border = NA)
polygon(c(hdi$ymax, dens$x[dens$x > hdi$ymax], rev(dens$x[dens$x > hdi$ymax])[1]),
        c(0, dens$y[dens$x > hdi$ymax], 0),
        col = "gray90", border = NA)

# Redraw density line on top
lines(dens, lwd = 3)
```




We can also visualise posterior samples as regression lines, to give a visual summary of the precision of model predictions

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 6
#| fig-height: 5.5

# basic plot
plot(d$distance, d$velocity, xlab = "Distance (Mpc)", ylab = "Velocity (km/s)", pch = 19)

# draw a subset of posterior lines
idx <- sample(seq_along(posterior_samples$beta), 500)
for(i in idx){
  abline(a = 0, b = posterior_samples$beta[i], lwd = 1, col = rgb(0,0,1,0.08))
}
abline(a = 0, b = median(posterior_samples$beta), lwd = 3, col = "blue")

# (redraw the points to prevent them being hidden by the lines)
points(d$distance, d$velocity, pch=19)
```


::: callout-tip

##### Using the "Generated quantities" block

Stan models have also other types of block. One of them is the generated quantities block, which we can use to compute quantities of interest that depends on posterior parameters. in this block we can include also functions that simulates data, so it is useful for things such as _posterior predictive checks_.

```{r, eval=FALSE}
stan_code <- "
data {
  int<lower=1> N;
  vector[N] distance; 
  vector[N] velocity; 
}

parameters {
  real<lower=0> beta; 
  real<lower=0> sigma;
}

model {
  beta ~ normal(0, 200);
  sigma ~ normal(0, 2000);
  velocity ~ normal(beta * distance, sigma);
}
generated quantities {
  real mpc_km; 
  real sec_per_year;

  mpc_km = 3.09 * 10^19;
  sec_per_year = 60^2 * 24 * 365;
  
  real age_universe = ((mpc_km / beta) / sec_per_year) / 1e9;
}
"

fit <- stan(
  model_code = stan_code,
  data = stan_data,
  iter = 2000,
  chains = 4)

print(fit, pars = c("beta","sigma", "age_universe"), probs = c(.025,.5,.975))
```

```{r,echo=FALSE}
fit <- readRDS("outputs/hubble_fit_2.RDS")
print(fit, pars = c("beta","sigma", "age_universe"), probs = c(.025,.5,.975))
```

:::

## Activity 3 — Overdispersed Poisson GLMM

> Goal: Model the rate of police stops across ethnic groups and precincts, adjusting for exposure (arrests) and allowing for *overdispersion* and *hierarchical structure*.


- **Context:**  Study of the NYPD *stop-and-frisk* policy, investigating possible ethnic bias in police stops.

- **Source:**  Records of ≈175,000 stops over 15 months (1998–1999), obtained by the New York State Attorney General’s Office.

- **Variables:**  
  - `stops` — Number of police stops recorded in 1998–1999 for the cell.
  - `pop` — Estimated population size of the ethnic group within the precinct (exposure candidate).
  - `past.arrests` — Number of 1997 arrests (DCJS) for the same group/crime category (proxy for crime exposure).
  - `precinct` — Precinct ID (integers **1–75**).
  - `eth` — Ethnicity code: **1 = Black**, **2 = Hispanic**, **3 = White**.
  - `crime` — Crime category: **1 = violent**, **2 = weapons**, **3 = property**, **4 = drug**.

#### Setup

Load data into R

```{r}
d <- read_delim("data/police_stops.txt")
str(d)
 print(d, n=14)
```

We first add crime and stop counts over crime type (i.e. we are pooling together all crime types)

```{r}
d <- d %>%
  group_by(precinct, eth) %>%
  summarise(stops = sum(stops),
         past.arrests = sum(past.arrests),
         pop = unique(pop))
```


Compute summary statistics (averaging across the whole city). 

```{r}
d_summaries <- d %>%
  mutate(eth = case_when(
    eth == 1 ~ "black",
    eth == 2 ~ "hispanic",
    eth == 3 ~ "white"
  )) %>%
  group_by(eth) %>%
  summarise(
    pop           = sum(pop),
    total_stops   = sum(stops),
    total_arrests = sum(past.arrests),
    .groups = "drop"
  ) %>%
  mutate(
    prop_of_all_stops = total_stops / sum(total_stops),
    pop_fraction      = pop / sum(pop)
  )

d_summaries
```

##### 1) Prepare data for Stan


Important: If `past.arrests == 0` for some rows but stops are nonzero, we need to avoid a zero Poisson mean. In this case however no adjustment is needed (otherwise we could have added a small constant, e.g. 0.5, to zeroes in `past.arrests`)

```{r}
any(d$past.arrests == 0)
```



```{r}
stan_data <- list(
  N         = nrow(d),
  n_eth     = length(unique(d$eth)),
  n_prec    = length(unique(d$precinct)),
  y         = d$stops,
  eth       = d$eth,
  past_arrests  = d$past.arrests,
  precinct  = d$precinct
)

str(stan_data)
```


##### 2) Stan code

The code of the model is in the file `overdispersed_poisson.stan`

```{.stan}

data {
  int<lower=1> N;          // total # of (eth, precinct) data points
  int<lower=1> n_eth;      // number of ethnicity categories, e.g. 3
  int<lower=1> n_prec;     // number of precincts
  
  int<lower=0> y[N];       // outcome counts y_{ep}
  vector[N] past_arrests;  // baseline/reference rate
  
  int<lower=1, upper=n_eth> eth[N];      // ethnicity ID for each row
  int<lower=1, upper=n_prec> precinct[N]; // precinct ID for each row
}

parameters {
  // Ethnicity effects (fixed)
  vector[n_eth] alpha; 
  
  // Random intercepts for precinct
  real<lower=0> sigma_beta; 
  vector[n_prec] beta_raw;   // non-centered param for precinct

  // Overdispersion
  real<lower=0> sigma_eps;
  vector[N] eps_raw;         // non-centered param for each (e,p) observation
}

transformed parameters {
  vector[n_prec] beta; 
  beta = sigma_beta * beta_raw;

  vector[N] eps;
  eps = sigma_eps * eps_raw;
}

model {
  // Priors
  alpha ~ normal(0, 5);           
  sigma_beta ~ exponential(1);    
  beta_raw ~ normal(0, 1);

  sigma_eps ~ exponential(1);     
  eps_raw ~ normal(0, 1);
  
  // Likelihood
  for (i in 1:N) {
    y[i] ~ poisson(past_arrests[i] * (15.0 / 12.0) * exp(alpha[eth[i]] + beta[precinct[i]] + eps[i]));
  }
}
```


Note that we use a *non-centred* parametrisation for the random intercepts:

$$\beta_p = \sigma_\beta \;\beta_{\text{raw},p}, \qquad \beta_{\text{raw},p} \sim N(0,1),$$

instead of sampling $\beta_p$ directly from $N(0,\sigma_\beta^2)$. This often improves sampling efficiency in hierarchical models. The same logic applies to \(\epsilon_{ep}\).


##### 3) Compile and run

```{r, eval=FALSE}
fit <- stan(
  file = "stan_code/overdispersed_poisson.stan",
  data = stan_data,
  iter = 2000,
  chains = 4,
  cores = 4)
```

```{r, echo=FALSE}
fit <- readRDS("outputs/police_stops.RDS")
```

We can use the print method for a quick look at the results

```{r}
print(fit, pars = c("alpha", "sigma_beta", "sigma_eps"))
```

Here, `alpha` are the ethnic-group log-effects, `sigma_beta` is the precinct-level standard deviation, and `sigma_eps` is the observation-level overdispersion parameter.

##### 4) Diagnostic checks

```{r}
# Traceplots
traceplot(fit, pars = c("alpha", "sigma_beta", "sigma_eps"))
```

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 5
#| fig-height: 5

# Pairs plot for some key parameters
pairs(fit, pars = c("alpha[1]", "alpha[2]", "alpha[3]", "sigma_beta", "sigma_eps"))
```

##### 5) Inference

We use `tidybayes` to extract samples and summarize. Suppose we want to look at the exponentiated ethnic effects (relative stops vs. arrests).

```{r}
library(tidybayes)

posterior <- fit %>% 
  spread_draws(alpha[e]) %>%
  mutate(rate = exp(alpha))  # exponentiate to interpret as a multiplicative factor

# Summarize the rate by ethnicity
posterior %>%
  group_by(e) %>%
  mean_hdi(rate, .width = 0.95)
```


If `e=1` = black, `e=2` = hispanic, `e=3` = white, these give the average ratio of stops to arrests for each ethnicity, accounting for precinct and overdispersion.

You might then compare black vs. white, etc., by calculating differences or ratios in the posterior draws.

```{r}
posterior_ratios <- posterior %>%
  pivot_wider(
    id_cols = c(.chain, .iteration, .draw),  # these 3 columns identify each draw
    names_from = e,                          # which column to pivot on
    values_from = rate,                      # which column holds the values
    names_prefix = "eth_") %>%
  mutate(
    black_vs_white = eth_1 / eth_3,
    hisp_vs_white  = eth_2 / eth_3) %>%
   # Pivot longer so each ratio is in its own row
  pivot_longer(
    cols = c(black_vs_white, hisp_vs_white),
    names_to = "contrast",
    values_to = "ratio"
  ) %>%
  # Summarize with mean_hdi -> returns columns .mean, .lower, .upper
  group_by(contrast) %>%
  mean_hdi(ratio, .width = 0.95)

posterior_ratios
```

This gives an estimate of how many times more likely stops of black (or hispanic) are relative to stops of white people, after controlling for arrests, precinct differences, and overdispersion.

To visualise this with a plot:

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 2.5
#| fig-height: 3

ggplot(posterior_ratios, aes(x = contrast, y = ratio)) +
  geom_point(size=3) +
  geom_hline(yintercept=1, lty=2)+
  geom_errorbar(aes(ymin = .lower, ymax = .upper), width = 0.1) +
  labs(
    x = NULL,
    y = "Ratio relative to White",
    title = "Ethnicity vs. White Stop Ratios",
    caption = "Points = posterior means; bars = 95% HDI"
  )
```


##### 5) Prior predictive check

To check that the prior is sensible, it can be useful to simulate the _prior predictive distribution_ and compare it to the actual, observed data. This is defined as 

$$p(y) =  \int p(y\mid \Theta) \ p( \Theta) d\Theta $$
where $\Theta$ indicate the parameters, in this case $\Theta = \left\{ \alpha, \sigma_{\beta}, \sigma_{\epsilon} \right\}$.

In other words, we resample both observation-level over dispersion values, the precinct-specific random intercepts, as well as the poisson observations. This gives us a way to estimate the prior predictive distribution, which we can then compare to the actual data. Ideally, the observed data should lie well within the prior-predictive distribution, suggesting that our priors were adequate and not too informative. 

```{r}
N <- nrow(d)
n_eth <- length(unique(d$eth))
n_prec <- length(unique(d$precinct))

# Decide how many draws from the prior to simulate
n_sims <- 500  

# We'll store all simulated Y values in a single big vector
# so we can compare them to the observed distribution
all_prior_stops <- numeric(0)

set.seed(42)  # reproducibility

for (s in 1:n_sims) {
  
  # 1) Sample the parameters from their priors
  alpha_draw <- rnorm(n_eth, mean = 0, sd = 5) # alpha: one for each ethnicity
  sigma_beta_draw <- rexp(1, rate = 1)
  beta_raw_draw <- rnorm(n_prec, 0, 1)
  beta_draw <- sigma_beta_draw * beta_raw_draw
  
  sigma_eps_draw <- rexp(1, rate = 1)          # Overdispersion
  eps_raw_draw <- rnorm(N, 0, 1)
  eps_draw <- sigma_eps_draw * eps_raw_draw
  
  # 2) Generate y (stops) from Poisson for each observation i
  past_arrests_offset <- ifelse(d$past.arrests == 0, 0.5, d$past.arrests)
  
  # compute the linear predictor for each row i
  lambda_vec <- numeric(N)
  for (i in seq_len(N)) {
    e <- d$eth[i]
    p <- d$precinct[i]
    lambda_vec[i] <- alpha_draw[e] + beta_draw[p] + eps_draw[i]
  }
  
  # Poisson means
  mu_vec <- past_arrests_offset * (15/12) * exp(lambda_vec)
  
  # sample from Poisson
  y_sim <- rpois(N, mu_vec)
  
  # 3) Store in a big vector for later comparison
  all_prior_stops <- c(all_prior_stops, y_sim)
}
```


Now we have the number of 'stops' simulated from the prior, and we can compare with the empirical distribution:

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 6
#| fig-height: 3
observed_stops <- d$stops

df_plot <- tibble(
  value = c((observed_stops ), (all_prior_stops )),
  type  = rep(c("Empirical", "Prior"), c(length(observed_stops), length(all_prior_stops))))

# Plot densities
ggplot(df_plot, aes(x = value, fill = type)) +
  geom_density(alpha = 0.4) +
  labs(x = "stops", y = "Density",
    title = "Prior Predictive vs. Empirical Distribution (log scale)",
    fill = "Data Source"
  ) +
  theme_minimal()+
  scale_x_log10()
```

##### 6) Posterior predictive check

We can use a similar approach to compare the posterior distribution to the observed data. That is, we are comparing the data with the _posterior predictive distribution_

$$p(y^{\text{rep}} \mid y) =  \int p(y^{\text{rep}} \mid \Theta) \ p( \Theta \mid y) d\Theta $$
It may be useful to not resample all of the random effects; in this case for example I decided to hold fixed the precinct-specific random intercepts, and to resample only the observation-level overdispersion values. 

This allows me to compute a predictive interval for the stop rate in each precinct, which allows more granularity in the comparison between model and data.

There are different ways to do this, either via R (as in the prior predictive check above) or we can take advantage of the generated quantities block in Stan, and add the following to our model code:

```{.stan}
generated quantities {
  vector[N] log_lik; // Log-likelihood for each observation (for LOO-CV)
  real eps_rep; // random effects observation-level overdipersion
  array[N] int y_rep;

  for (i in 1:N) {

    // Compute log-likelihood for observed data    
    log_lik[i] = poisson_lpmf(y[i] | past_arrests[i] * (15.0 / 12.0) * exp(alpha[eth[i]] + beta[precinct[i]] + eps[i]));

    // Simulate replicated data from posterior predictive distribution
    eps_rep = sigma_eps * normal_rng(0, 1);
    y_rep[i] = poisson_rng(past_arrests[i] * (15.0 / 12.0) * exp(alpha[eth[i]] + beta[precinct[i]] + eps_rep));
  }
}

```

Assuming the code is saved in the file `stan_code/overdispersed_poisson_PPC.stan`, we then need to re-do the sampling:

```{r, eval=FALSE}
fit <- stan(
  file = "stan_code/overdispersed_poisson_PPC.stan",
  data = stan_data,
  iter = 2000,
  chains = 4,
  cores = 4)

```
```{r, echo=FALSE}
fit <- readRDS("outputs/police_stops_PPC.RDS")
```

Once we have run that, we can extract the posterior predictive for each precinct, and plot them alongside the observed.


```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 5
#| fig-height: 8

# Map observation index -> precinct
precinct_idx <- d$precinct  # length N

# Observed precinct totals
obs_totals <- d %>%
  group_by(precinct) %>%
  summarise(obs_total = sum(stops), .groups = "drop")

# Exposure per precinct (for *rates* option)
exposure_by_precinct <- d %>%
  mutate(expo = past.arrests * (15/12)) %>%
  group_by(precinct) %>%
  summarise(exposure = sum(expo), .groups = "drop")

# 1) Extract posterior predictive draws for y_rep[i]
yrep_long <- fit %>%
  spread_draws(y_rep[i]) %>%
  mutate(precinct = precinct_idx[i])

# 2) Aggregate to precinct totals per draw
ppc_totals <- yrep_long %>%
  group_by(.draw, precinct) %>%
  summarise(rep_total = sum(y_rep), .groups = "drop")

# 3a) (Totals) Summarise mean and 95% CI per precinct
ppc_totals_sum <- ppc_totals %>%
  group_by(precinct) %>%
  mean_qi(rep_total, .width = 0.95) %>%    # columns: rep_total, .lower, .upper
  ungroup() %>%
  left_join(obs_totals, by = "precinct")

ppc_totals_sum_ord <- ppc_totals_sum %>%
  mutate(precinct_ord = fct_reorder(factor(precinct), obs_total))


# 4a) Plot totals: posterior predictive (mean + 95% CI) vs observed
ppc_totals_sum %>%
  mutate(precinct = fct_reorder(factor(precinct), obs_total)) %>%
  ggplot(aes(x = factor(precinct), y = rep_total)) +
  geom_pointrange(aes(ymin = .lower, ymax = .upper), color="dark grey") +
  geom_point(aes(y = obs_total), shape = 4, size = 2, position = position_nudge(x = 0.15)) +
  labs(x = "Precinct", y = "Total stops",
       title = "Posterior predictive totals by precinct",
       subtitle = "Points = posterior mean (CI = 95%), X = observed total") +
  coord_flip() +
  theme_minimal()
```

## Activity 4 —  Lapse-augmented probit GLMM

> Goal: Model binary perceptual choices (left/right) as a function of signed distance, allowing precision and bias to vary across conditions and observers, and incorporating an observer-specific lapse rate to capture random guessing.

#### Setup

Load data into R

```{r}
d <- read.table("data/bisection2.txt", sep="\t",header=T)
str(d)
```

#### Visualisation


```{r}
#| echo: FALSE
#| fig-height: 4
#| fig-width: 11
#| fig-align: center

binomSEM <- function (v) {
    sqrt((mean(v) * (1 - mean(v)))/length(v))}

d$dx_bin <- cut(d$dx, 15)
dag <- aggregate(cbind(rr,dx)~id+dx_bin+cond, d, mean)
dag$se <- aggregate(rr~id+dx_bin+cond, d, binomSEM)$rr
dag$n <- aggregate(rr~id+dx_bin+cond, d, length)$rr

ggplot(dag, aes(x=dx,y=rr,color=factor(cond)))+
  geom_errorbar(aes(ymin=rr-se,ymax=rr+se),width=0)+
  geom_point(pch=21)+
  facet_wrap(~id,ncol=7)+
  theme_minimal()+
  geom_smooth(data=d,aes(x=dx,y=rr),
              method="glm",
              method.args=list(family=binomial(probit)),
              size=0.8,fullrange=T,se=F)+
  scale_color_manual(values=c("black","blue"),guide=F)+
  labs(y="choice probability", x="distance offset [deg]")
```


##### 1) Prepare data for Stan


```{r}
d_stan <- list(
  id=as.numeric(factor(d$id, labels=1:length(unique(d$id)))), 
  ds=d$dx, 
  rr = d$rr, 
  N=nrow(d), 
  J=length(unique(d$id)), 
  cond=d$cond-1)

str(d_stan)
```


##### 2) Stan code

Model code available in `stan_code/psyGLMM.stan`


```{.stan}
data {
  int<lower=1> N;
  array[N] int<lower=0, upper=1> rr;
  array[N] real ds;
  array[N] int<lower=0,upper=1> cond;
  int<lower=1> J;
  array[N] int<lower=1,upper=J> id;
}

parameters {
  vector[4] beta;                    // fixed-effects parameters
  vector<lower=0,upper=1>[J] lambda; // lapse rate
  vector<lower=0>[4] sigma_u;        // random effects standard deviations
  cholesky_factor_corr[4] L_u;       // L_u is the Choleski factor of the correlation matrix
  matrix[4,J] z_u;                   // random effect matrix
}

transformed parameters {
  matrix[4,J] u;
  u = diag_pre_multiply(sigma_u, L_u) * z_u; // use Cholesky to set correlation
}

model {
  real mu; // linear predictor

  //priors
  L_u ~ lkj_corr_cholesky(2);   // LKJ prior for the correlation matrix
  to_vector(z_u) ~ normal(0,1); // before Cholesky, random effects are normal variates with SD=1
  sigma_u ~ cauchy(0, 1);       // SD of random effects (vectorized)
  beta ~ normal(0, 2);          // prior 
  lambda ~ beta(1, 7);          // lapse rate ~50% prior that <0.1, pbeta(0.1, 1, 7)

  //likelihood
  for (i in 1:N){
    
    mu = beta[1] + u[1,id[i]] + 
      cond[i]*(beta[2]+u[2,id[i]]) + 
      (beta[3] + u[3,id[i]] + cond[i]*(beta[4] + u[4,id[i]] ))*ds[i];
    
    rr[i] ~ bernoulli((1-lambda[id[i]])*Phi(mu)+lambda[id[i]]/2);
  }
}
```

##### 3) Compile and run

```{r, eval=FALSE}
fit <- stan(file = "stan_code/psyGLMM.stan", data = d_stan, iter = 4000, chains = 4)
```

```{r, echo=FALSE}
#saveRDS(fit,file="outputs/psyGLMM.RDS")
fit <- readRDS("outputs/psyGLMM.RDS")

```

We can use the usual functions to check that the models has converged

```{r}
traceplot(fit, pars = c("beta", "sigma_u"))
```

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 7
#| fig-height: 7

# Pairs plot for some key parameters
pairs(fit, pars = c("beta", "sigma_u"))
```

```{r}
print(fit, pars = c("beta", "sigma_u"), 
    probs = c(0.025, 0.975), digits = 3)
```

##### 4) Inference

The package `tidybayes` has useful function to visualise the posterior distribution of model parameters:

```{r}
#| fig-align: 'center'
#| fig-cap-location: margin
#| fig-width: 6
#| fig-height: 5

fit %>%
  gather_draws(beta[k], sigma_u[r]) %>%
  mutate(term = if_else(.variable == "beta",
                        paste0("beta[", k, "]"),
                        paste0("sigma_u[", r, "]"))) %>%
  mutate(term = factor(term, levels = c(paste0("beta[", 1:4, "]"),
                                        paste0("sigma_u[", 1:4, "]")))) %>%
  ggplot(aes(x = .value, y = fct_rev(term))) +
  geom_vline(xintercept=0, lty=2, size=0.4) +
  stat_halfeye(.width = 0.8, normalize = "groups") +
  labs(x = "Posterior value", y = NULL,
       title = "Posterior densities (80% credible intervals): β and σ") +
  theme_minimal()
```


Rather than GLMM coefficients, the measures of interests are the mean and SD of the cumulative Gaussian function used to model choice probabilities. The coefficients of the GLM can be straighforwardly mapped to the parameters of the psychometric function $$
\begin{align}
p \left(y_i\right) & = \Phi\left(\frac{x_i -\mu}{\sigma} \right) \\
& = \Phi ( -\frac{\mu}{\sigma} + \frac{1}{\sigma}x_i )\\
& = \Phi\left(\beta_0 + \beta_1x_i \right) \\ \\& \mu=-\frac{\beta_0}{\beta_1}, \,\,\, \sigma=\frac{1}{\beta_1} \\
\end{align}
$$
We can use these relationships to estimate a Bayesian CI around the difference in the slope of the functions across the two conditions:


```{r}
spread_draws(fit, beta[k]) %>%
  mutate(beta = set_names(beta, k)) %>%
  pivot_wider(names_from = k, values_from = beta, names_prefix = "beta") %>%
  transmute(
    JND_1    = 1 / beta3,
    JND_2    = 1 / (beta3 + beta4),
    JND_diff = (1 / (beta3 + beta4)) - (1 / beta3)
  ) %>%
  pivot_longer(everything(), names_to = "quantity", values_to = "value") %>%
  group_by(quantity) %>%
  mean_hdi(value, .width = 0.95)
```


